---
title: Estimating spatial processes in a simulation context with TMB.
author: Kelli Faye Johnson
output: pdf_document
header-includes:
  - \usepackage{bm}
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

```{r setup, echo = FALSE, results = "hide", , warning = FALSE, message = FALSE}

knitr::opts_chunk$set(fig.width = 7, fig.height = 9.5,
  echo = FALSE, warning = FALSE, message = FALSE)

library(ggplot2)
library(INLA)
library(TMB)
cpp <- "spatial_gompertz"
compile(file.path("data", paste0(cpp, ".cpp")))
dyn.load(TMB::dynlib(file.path("data", cpp)))

# source the library of functions for this repo
source("alaska_ggplot_theme.R")
ignore <- sapply(dir("lib", full.names = TRUE), source)

# Functions
fmt <- function(x) {
  format(x, nsmall = 2, scientific = FALSE, digits = 2)
}
```

```{r simulationinputs, results = "hide"}
n_replicates <- 100
variables <- c("Simulated_example", "zeroinflatedlnorm")
n_years <- 10
n_stations <- 100
scale <- 0.25
SD_O <- 0.4
SD_E <- SD_O / 2
SD_obs <- SD_O / 4
log_clustersize <- log(4)
rho <- 0.5
logMeanDens <- 1.0
models <- c("RMgauss", "RMgneiting")
multiplier <- 5

savedin <- paste0("alaska_simulation_", n_stations)
  newname <- strsplit(savedin, "_")[[1]]
  newname[grepl("[0-9]+", newname)] <- as.numeric(newname[grepl("[0-9]+", newname)]) * multiplier
  newname <- paste(newname, collapse = "_")
dir.create(savedin, showWarnings = FALSE)
dir.create(newname, showWarnings = FALSE)
```

# Introduction
Template Model Builder (TMB) offers a framework to estimate random effects.
Here, we are concerned with the ability of TMB to estimate
spatially explicit random effects through the use of sparse covariance
matrices, or Gaussian Markov Random Fields (GMRFs).

We perform a simulation experiment where spatially referenced count data
were sampled from a true population generated using the
`RandomFields` package in **R** and fit to
a spatially-explicit Gompertz population dynamics model in TMB.
Specific parameterizations of the
operating model (OM; generator of the truth)
and estimation method (EM) may or may not affect the
the ability of TMB to estimate parameters of importance without bias.

# Methods

The true population was governed by a spatially-explicit Gompertz population
dynamics model,
$$
log(\bm{n}_{t}) = \bm{\omega} + \rho * log(\bm{n}_{t-1}) + \bm{\epsilon}_{t},
$$
where $\bm{n}_{t}$ a vector representing the density of the population at each
sampled location in time $t$, $\bm{\omega}$ is a vector of spatially-explicit deviates in
productivity from the mean productivity, $\alpha$, $\rho$ represents the degree of
density dependence, and $\bm{\epsilon}_{t}$ is a vector of process error in year $t$.
The initial conditions (i.e., $t=1$) were specified as,
$$
log(\bm{n}_{1}) = \frac{\bm{\omega}}{1 - \rho} + \phi\bm{1} + \bm{\epsilon}_{1},
$$
where $\phi\bm{1}$ is the log-ratio of expected abundance in the initial year and the
median of the stationary distribution at equilibrium.
Stochastic processes were assumed to follow multivariate normal distributions,
$\bm{\omega} \sim MVN(\alpha\bm{1}, \sigma^2_{\omega}\bm{R})$ and
$\bm{\epsilon_{t}} \sim MVN(0, \sigma^2_{\epsilon}\bm{R})$.
Marginal standard deviations were assumed to be independent, but the Mat&eacute;rn
spatial correlation function, $\bm{R}$, that approximates the stochastic process as a
GRF was assumed to be the same for productivity and process error.

Each simulation included `r n_years` years of data from `r n_stations` unique
sampling locations, where all sampling locations were sampled on a yearly basis,
repeated `r n_replicates` times.
The capabilities of the EM was assessed using relative error (RE)
and median absolute relative error (MARE).
RE was calculated using
$\frac{(true - estimate)}{true}$
and
$\frac{(e^{true} - e^{estimate})}{e^{true}}$ for parameters where
the true value was non-zero and zero, respectively. The latter allows
for non-infinite RE values.

```{r simulatedata, results = "hide", cache = TRUE}
sim_data <- NULL
for (it_rep in seq(n_replicates)) {
for (it_model in models) {
  sim_data <- rbind(sim_data, data.frame(Sim_Gompertz_Fn(
    n_years = n_years,
    n_stations = n_stations * multiplier,
    phi = 0.0,
    rho = rho,
    logMeanDens = logMeanDens,
    SpatialScale = scale,
    SD_O = SD_O,
    SD_E = SD_E,
    SD_obs = SD_obs,
    log_clustersize = log_clustersize,
    model = it_model,
    seed = it_rep + 10),
    "replicate" = it_rep,
    "subpopulations" = 1,
    "percentinc" = 1))
  lots_data <- sim_data
  sim_data <- sim_data[sim_data$Site %in% 1:n_stations, ]
}}
```

```{r estimate, results = "hide", eval = FALSE}
for (it_rep in seq(n_replicates)) {
for (it_model in models) {
for (it_variable in variables) {
  set.seed(it_rep + 10)

  data_sub <- sim_data[sim_data$replicate == it_rep & sim_data$model == it_model, ]
  # Build SPDE object using INLA
  mesh <- INLA::inla.mesh.create(data_sub[, c("Longitude", "Latitude")],
    boundary = NULL, plot.delay = NULL, cutoff = 1e-12,
    extend = list(n = 8, offset = -0.15),
    refine = list(min.angle = 26))

  # Make object
  obj <- calc_adfun(data = data_sub, mesh = mesh, tmb = cpp,
    variable = it_variable)

  # Run optimizer
  opt3 <- nlminb(obj$par, objective = obj$fn,
    gradient = obj$gr,
    lower = c(rep(-200, 5), -0.999, rep(-200, 2)),
    upper = c(rep( 200, 5), 0.999, rep(200, 2)),
    control = list(eval.max = 1e4, iter.max = 1e4, trace = 1))

  # Get standard errors
  Report <- obj$report()
  Report$gradient <- obj$gr(opt3$par)
  Report$SD3 <- try(sdreport(obj))

  Report$DF <- data_sub
  Report$variable <- it_variable
  Report$mesh <- mesh
  Report$file <- file.path(savedin,
    paste0("sim-", it_model, "-", it_variable, "-", it_rep, ".RData"))

  save(Report, file = Report$file)
  rm(data_sub, mesh, obj, opt3, Report)
}}}
```

```{r lotsofdata, results = "hide", eval = FALSE}
for (it_rep in seq(n_replicates)) {
for (it_model in models) {
for (it_variable in variables) {
  set.seed(it_rep + 10)

  data_sub <- lots_data[sim_data$replicate == it_rep & sim_data$model == it_model, ]
  # Build SPDE object using INLA
  mesh <- INLA::inla.mesh.create(data_sub[, c("Longitude", "Latitude")],
    boundary = NULL, plot.delay = NULL, cutoff = 1e-12,
    extend = list(n = 8, offset = -0.15),
    refine = list(min.angle = 26))

  # Make object
  obj <- calc_adfun(data = data_sub, mesh = mesh, tmb = cpp,
    variable = it_variable)

  # Run optimizer
  opt3 <- nlminb(obj$par, objective = obj$fn,
    gradient = obj$gr,
    lower = c(rep(-200, 5), -0.999, rep(-200, 2)),
    upper = c(rep( 200, 5), 0.999, rep(200, 2)),
    control = list(eval.max = 1e4, iter.max = 1e4, trace = 1))

  # Get standard errors
  Report <- obj$report()
  Report$gradient <- obj$gr(opt3$par)
  Report$SD3 <- try(sdreport(obj))

  Report$DF <- data_sub
  Report$variable <- it_variable
  Report$mesh <- mesh
  Report$file <- file.path(newname,
    paste0("sim-", it_model, "-", it_variable, "-", it_rep, ".RData"))

  save(Report, file = Report$file)
  rm(data_sub, mesh, obj, opt3, Report)
}}}
```

### Sensitivity analyses

Two sensitivity analyses were run, one where $\alpha$ was fixed at its true value and
the one where $\rho$ was fixed at its true value.
These parameters were chosen because initial exploration indicated
that they exhibited the highest correlation with the remaining model parameters,
 mainly with the marginal standard deviation of $\bm{\omega}$, $\sigma_{\omega}$.

```{r sensitivity_fixalpha, results = "hide", eval = FALSE}

dir.create(file.path(savedin, "fixed_alpha"), showWarnings = FALSE)
it_model <- models[1]
it_variable <- variables[1]
for (it_rep in seq(n_replicates)) {
  set.seed(it_rep + 10)

  data_sub <- sim_data[sim_data$replicate == it_rep & sim_data$model == it_model, ]
  # Build SPDE object using INLA
  mesh <- INLA::inla.mesh.create(data_sub[, c("Longitude", "Latitude")],
    boundary = NULL, plot.delay = NULL, cutoff = 1e-12,
    extend = list(n = 8, offset = -0.15),
    refine = list(min.angle = 26))

  # Make object
  obj <- calc_adfun(data = data_sub, mesh = mesh, tmb = cpp,
    variable = it_variable, fixed = "alpha")

  # Run optimizer
  opt3 <- nlminb(obj$par, objective = obj$fn,
    gradient = obj$gr,
    control = list(eval.max = 1e4, iter.max = 1e4, trace = 1))

  # Get standard errors
  Report <- obj$report()
  Report$gradient <- obj$gr(opt3$par)
  Report$SD3 <- try(sdreport(obj))

  Report$DF <- data_sub
  Report$variable <- it_variable
  Report$mesh <- mesh
  Report$file <- file.path(savedin, "fixed_alpha",
    paste0("sim-", it_model, "-", it_variable, "-", it_rep, ".RData"))

  save(Report, file = Report$file)
  rm(data_sub, mesh, obj, opt3, Report)
}
```

```{r sensitivity_fixrho, eval = FALSE, results = "hide"}

dir.create(file.path(savedin, "fixed_rho"), showWarnings = FALSE)
it_model <- models[1]
it_variable <- variables[1]
for (it_rep in seq(n_replicates)) {
  set.seed(it_rep + 10)

  data_sub <- sim_data[sim_data$replicate == it_rep & sim_data$model == it_model, ]
  # Build SPDE object using INLA
  mesh <- INLA::inla.mesh.create(data_sub[, c("Longitude", "Latitude")],
    boundary = NULL, plot.delay = NULL, cutoff = 1e-12,
    extend = list(n = 8, offset = -0.15),
    refine = list(min.angle = 26))

  # Make object
  obj <- calc_adfun(data = data_sub, mesh = mesh, tmb = cpp,
    variable = it_variable, fixed = "rho")

  # Run optimizer
  opt3 <- nlminb(obj$par, objective = obj$fn,
    gradient = obj$gr,
    control = list(eval.max = 1e4, iter.max = 1e4, trace = 1))

  # Get standard errors
  Report <- obj$report()
  Report$gradient <- obj$gr(opt3$par)
  Report$SD3 <- try(sdreport(obj))

  Report$DF <- data_sub
  Report$variable <- it_variable
  Report$mesh <- mesh
  Report$file <- file.path(savedin, "fixed_rho",
    paste0("sim-", it_model, "-", it_variable, "-", it_rep, ".RData"))

  save(Report, file = Report$file)
  rm(data_sub, mesh, obj, opt3, Report)
}
```

# Results

In general, the EM can estimate the marginal variation in process error,
the scale of the spatial effects, and $\phi$,
but not the remaining parameters (Figures 1 and 2).
The Poisson and Poisson lognormal ('Poissonln') objective functions
performed similarly (Figure 1), namely because the model was able to estimate both
of the additional parameters governing the Poissonln function (Figure 3).
Additionally, there was no appreciable difference between the two available functions in
the `RandomFields` package used to simulate MVN processes (i.e., `RMgauss` and `RMgneiting`; where the latter is supposed to be less biased).

Parameter estimates improved when either $\alpha$ (Figure 4) or $\rho$ (Figure 5)
were fixed at their true values.

```{r readinresults, echo = FALSE, results = "hide", cache = FALSE}
resultfiles <- dir(savedin, full.names = TRUE, pattern = "\\.RData",
  recursive = FALSE, include.dirs = FALSE)
all <- list()
for (it_ in resultfiles) {
  all[[length(all) + 1]] <- read_results(file = it_, dir = NULL, projection = NULL)
}
all <- do.call("rbind", all)
all <- calc_RE(data = all, return = "all")
```

```{r readinresults_lotsdata, results = "hide", cache = FALSE}
resultfiles <- dir(newname, full.names = TRUE, pattern = "\\.RData",
  recursive = FALSE, include.dirs = FALSE)
lots_data_results <- list()
for (it_ in resultfiles) {
  lots_data_results[[length(lots_data_results) + 1]] <- read_results(file = it_, dir = NULL, projection = NULL)
}
lots_data_results <- do.call("rbind", lots_data_results)
lots_data_results <- calc_RE(data = lots_data_results, return = "all")
```

```{r readinresults_sensitivity, results = "hide", cache = FALSE}
resultfiles <- dir(file.path(savedin, "fixed_alpha"), full.names = TRUE,
  pattern = "\\.RData", recursive = TRUE, include.dirs = FALSE)
fixed_alpha <- list()
if (length(resultfiles) > 0) {
  for (it_ in resultfiles) {
    fixed_alpha[[length(fixed_alpha) + 1]] <- read_results(file = it_, dir = NULL, projection = NULL)
    # all[[length(all)]] <- calc_cluster(all[[length(all) + 1]], weight = 0.1)
  }
  fixed_alpha <- do.call("rbind", fixed_alpha)
  fixed_alpha <- calc_RE(data = fixed_alpha, return = "all")
}

resultfiles <- dir(file.path(savedin, "fixed_rho"), full.names = TRUE,
  pattern = "\\.RData", recursive = TRUE, include.dirs = FALSE)
fixed_rho <- list()
if (length(resultfiles) > 0) {
  for (it_ in resultfiles) {
    fixed_rho[[length(fixed_rho) + 1]] <- read_results(file = it_, dir = NULL, projection = NULL)
    # all[[length(all)]] <- calc_cluster(all[[length(all) + 1]], weight = 0.1)
  }
  fixed_rho <- do.call("rbind", fixed_rho)
  fixed_rho <- calc_RE(data = fixed_rho, return = "all")
}
```

# Figures

```{r plot_iteration1, eval = FALSE, include = FALSE, fig.height = 4, fig.width = 8, fig.cap = "Sampled counts at each station for the first replicate."}
## Plot iteration 1
 ggplot(data = sim_data[sim_data$replicate == 1, ],
   aes(x = Year, y = Simulated_example, group = Site)) +
 ylab("Simulated count per location") +
 geom_line(alpha = 0.35) +
 theme(
   plot.background = element_blank(),
   panel.grid.major = element_blank(),
   panel.grid.minor = element_blank(),
   strip.background = element_blank(),
   panel.background = element_blank(),
   panel.border = element_rect(colour = "white", fill = NA, size = 1),
   legend.key = element_rect(colour = "white"),
   legend.title = element_text(size = 0, face = "bold"),
   legend.text = element_text(size = 7, face = "bold"))
```

## Sample size of `r n_stations` per year

```{r fig_pars, fig.height = 8, fig.cap = "Parameter estimates and their median relative error and median absolute relative error. Red dashed line indicates the true parameter value."}
plot_pars(all,
  pars = c("alpha", "sigma[epsilon]", "sigma[omega]", "scale", "rho"))
```

```{r fig_pars_box}
plot_pars(all,
  pars = c("alpha", "sigma[epsilon]", "sigma[omega]", "scale", "rho"),
  type = "boxplot")
```

```{r fig_phi, echo = FALSE, fig.cap = "Estimates of $\\phi$ and its median relative error (RE) and median absolute RE. Red dashed line indicates the true parameter value. RE was calculated using exponentiated values to circumvent RE values of infinity."}
plot_pars(all, pars = c("phi"))
```

```{r fig_lnPoisson, echo = FALSE, fig.cap = "Parameter estimates and their median relative error and median absolute relative error for the estimation method that assumed a Poisson lognormal response variable. Red dashed line indicates the true parameter value."}
plot_pars(all[all$variable == "zeroinflatedlnorm", ],
  pars = c("ln(size)", "sigma[obs]"))
```

```{r fig_fixed_alpha, eval = TRUE, fig.cap = "Parameter estimates and their median absolute relative error when $\\alpha$ was fixed at its true value. Red dashed line indicates the true parameter value."}
plot_pars(fixed_alpha,
  pars = c("rho", "sigma[epsilon]", "sigma[omega]", "scale"))
```

```{r fig_fixed_rho, eval = TRUE, fig.cap = "Parameter estimates and their median absolute relative error when $\\rho$ was fixed at its true value. Red dashed line indicates the true parameter value."}
plot_pars(fixed_rho,
  pars = c("alpha", "sigma[epsilon]", "sigma[omega]", "scale"))
```

## Sample size of `r n_stations * multiplier` per year

```{r fig_lotsdata_pars, fig.height = 8, fig.cap = "Parameter estimates and their median relative error and median absolute relative error. Red dashed line indicates the true parameter value."}
plot_pars(lots_data_results,
  pars = c("alpha", "sigma[epsilon]", "sigma[omega]", "scale", "rho"))
```

```{r fig_lots_data_results_pars_box}
plot_pars(lots_data_results,
  pars = c("alpha", "sigma[epsilon]", "sigma[omega]", "scale", "rho"),
  type = "boxplot")
```

```{r fig_lots_data_results_phi, echo = FALSE, fig.cap = "Estimates of $\\phi$ and its median relative error (RE) and median absolute RE. Red dashed line indicates the true parameter value. RE was calculated using exponentiated values to circumvent RE values of infinity."}

plot_pars(lots_data_results, pars = c("phi"))
```

```{r fig_lots_data_results_lnPoisson, echo = FALSE, fig.cap = "Parameter estimates and their median relative error and median absolute relative error for the estimation method that assumed a Poisson lognormal response variable. Red dashed line indicates the true parameter value."}
plot_pars(lots_data_results[lots_data_results$variable == "zeroinflatedlnorm", ],
  pars = c("ln(size)", "sigma[obs]"))
```

```{r manunit, eval = FALSE, echo = FALSE, message = FALSE, warning = FALSE}
###############################################################################
# Plot management unit estimates for 2 subpopulations
###############################################################################
# png(file.path(dir.results, "simulation_scen2_manunitboxes.png"),
#   width = my.width[2], units = "in", res = my.resolution, height = my.width[2])

# temp <- aggregate(Longitude ~
#   percentinc + n_years + replicate + group,
#   data = all[all$Year == 1, ], length)
# colnames(temp)[which(colnames(temp) == "Longitude")] <- "true"
# temp <- merge(temp,
#   aggregate(Longitude ~
#   percentinc + n_years + replicate + group + cluster,
#   data = all[all$Year == 1, ], length),
#   all = TRUE)
# temp$percent <- with(temp, Longitude / true)
# temp$percentinc <- factor(temp$percentinc,
#   labels = c("baseline", unique(temp$percentinc)[-1]))
# temp$percentinc[temp$percentinc == 0] <- "baseline"
# temp$cluster <- factor(temp$cluster, levels = unique(temp$cluster))
# temp$percent[which(temp$group != temp$cluster)] <-
#   temp$percent[which(temp$group != temp$cluster)] * -1

# ggplot(data = temp,
#   aes(x = cluster, y = percent)) +
#   geom_boxplot() +
#   facet_grid(percentinc + n_years ~ group, labeller = label_parsed) +
#   xlab("management unit") + ylab("proportion of subpopulation in management unit") +
#   labs(title = "subpopulation") +
#   theme(legend.position = "none", axis.title = element_text(size = 13),
#     plot.title = element_text(size = 13)) +
#   my.theme +
#   geom_hline(yintercept = 0, lty = 2, col = "red") +
# geom_rect(data = temp[temp$group == temp$cluster, ],
#     aes(xmin = as.numeric(cluster) - 0.5, xmax = as.numeric(cluster) + 0.5,
#     ymin = -1.1, ymax = 1.1),
#     size = 0.5, alpha = 0, color = "red", lty = 2)
# dev.off()
# rm(temp)
# }
```
