---
title: RandomFields, an R package, and properties of simulated data using it.
author: Kelli Faye Johnson
output:
  pdf_document:
    fig_caption: yes
date: "2017-01-19"
---

```{r setup, echo = FALSE, eval = TRUE, message = FALSE, include = FALSE}
library(INLA)
library(ncf)
library(RandomFields)
library(TMB)
ignore <- sapply(dir("lib", full.names = TRUE), source)
variance <- 0.5
scale <- 0.25
phi <- 6
reps <- 10
n <- 1000
scale_small <- 0.001
compile(file.path("data", "test.cpp"))
dyn.load(dynlib(file.path("data", "test")))
```

All data was simulated using the `RandomFields` package in `R`.
Whereas, estimation was done within the TMB framework, so there may be
miss-specifications that cannot be attributed to estimation capabilities, but
rather just due to differences in frameworks. For instance,
the scale (operating model) and range (estimation method) are not equivalent, and
differ by a scalar of two.

### No spatial correlation
When scale was set to a small number (i.e., `r scale_small`) the spatial correlation
in the simulated data was negligible and standard functions could be used
to calculate the realized standard deviation of the simulated vectors of
spatial deviation (e.g., $\omega_{x,y}$).
Scale, which represents the distance at which the correlation falls to
0.10 of its maximum, cannot be set to zero, and any value smaller than
`r scale_small` was sufficient to remove spatial autocorrelation in the
simulated data.
```{r noscale, echo = FALSE}
  sim_data <- Sim_Gompertz_Fn(
    n_years = 20,
    n_stations = n,
    SpatialScale = scale_small,
    SD_O = sqrt(variance),
    SD_E = sqrt(variance),
    SD_obs = 0.1,
    logMeanDens = 1.0,
    rho = 0.5,
    phi = 0.0,
    Loc = NULL,
    projection = NULL,
    seed = 5)
```

The true variance used in the call to `RandomFields` was `r variance`,
or `sd = `r round(sqrt(variance), 2)``.

```{r boxplotsdsmall, echo = FALSE, eval = TRUE, fig.cap = "Estimated standard deviation of the simulated spatial fields using RandomFields package with a small scale value."}
ignore <- data.frame("re" = c("sigma[omega]",
  rep("sigma[epsilon]", length(unique(sim_data$Year)))),
  "sigma" = c(sd(sim_data$Omega[sim_data$Year == 1]),
  tapply(sim_data$Epsilon, sim_data$Year, sd)))
foo <- Vectorize(function(u) eval(parse(text=sprintf("expression(%s)", u))))
b <- boxplot(sigma ~ re, data = ignore, plot = FALSE)
boxplot(sigma ~ re, data = ignore, names = foo(b$names),
  ylab = expression(sigma), las = 1)
abline(h = sqrt(variance), lty = 2, col = "red")
text(x = 1:2, y = tapply(ignore$sigma, ignore$re, mean) * 0.99,
  labels = paste0("(n = ", table(ignore$re), ")"))
```

### Spatial correlation
When the scale value was set to `r scale`, the base function in `R` to
calculate standard deviation no longer gave the standard deviation used
to generate the data.

```{r smallscale, echo = FALSE}
  sim_data <- Sim_Gompertz_Fn(
    n_years = 20,
    n_stations = n,
    SpatialScale = scale,
    SD_O = sqrt(variance),
    SD_E = sqrt(variance),
    SD_obs = 0.1,
    logMeanDens = 1.0,
    rho = 0.5,
    phi = 0.0,
    Loc = NULL,
    projection = NULL,
    seed = 5)
```

```{r boxplotsdlarge, echo = FALSE, fig.cap = "Estimated standard deviation of the simulated spatial fields using RandomFields package with a significant scale value."}
ignore <- data.frame("re" = c("sigma[omega]",
  rep("sigma[epsilon]", length(unique(sim_data$Year)))),
  "sigma" = c(sd(sim_data$Omega[sim_data$Year == 1]),
  tapply(sim_data$Epsilon, sim_data$Year, sd)))
b <- boxplot(sigma ~ re, data = ignore, plot = FALSE)
boxplot(sigma ~ re, data = ignore, xlab = "Random effect",
 names = foo(b$names), ylab = expression(sigma))
abline(h = sqrt(variance), lty = 2, col = "red")
text(x = 1:2, y = tapply(ignore$sigma, ignore$re, mean) * 0.99,
  labels = paste0("(n = ", table(ignore$re), ")"))
```

Though the Moran's correlation gave back the true standard deviation
at half of the input scale value.

```{r moran, echo = FALSE, fig.cap = "Moran's correlation versus distance (m)."}
par(mfrow = c(2, 1), oma = c(4, 4, 0.5, 0.5), mar = rep(0, 4),
  las = 1)
for (a in c("Omega", "Epsilon")) {
  test <- with(sim_data[sim_data$Year == 1, ],
    ncf::correlog(Longitude, Latitude,
    eval(parse(text = a)), increment = scale/100, resamp = 100,
    quiet = TRUE))
  with(test, plot(mean.of.class, correlation,
    xlim = c(0, scale * 1.5),
    ylim = c(0, sqrt(variance) * 1.5),
    xaxt = ifelse(a == "Omega", "n", "s")))
  abline(v = scale/2, lty = 2)
  abline(h = sqrt(variance), lty = 2)
  mtext(side = 3, line = -3, text = a, adj = 1)
}
par(las = 0)
mtext(side = 1, text = "Distance", line = 2, outer = TRUE)
mtext(side = 2, text = "Moran correlation", line = 2.5, outer = TRUE)
```

### Distributions
Code was originally adapted from that written by Dr. James T. Thorson, and, thus,
the `RandomFields::RMgauss` function was used to generate a
gaussian markov random field (GMRF).
The documentation for
[`RandomFields`](https://cran.r-project.org/web/packages/RandomFields/RandomFields.pdf)
specifies that `RandomFields::RMgneiting` is hardly distinguishable from
`RandomFields::RMgneiting`, but possesses neither the mathematical nor the numerical
disadvantages of the Gaussian model.

The following code will show their similarity. Unfortunately, the graphic device is not
compatible with `rmarkdown`.
```{r similar, eval = FALSE}
plot(RandomFields::RMgauss(var = variance, scale = scale))
lines(RandomFields::RMgneiting(orig = FALSE, var = variance, scale = scale),
  lty = 2)
legend("topright", bty = "n", legend = c("Gaussian", "Gneiting"),
  lty = c(1:2))
```

```{r cleanup, echo = FALSE, eval = TRUE, message = FALSE, include = FALSE}
dyn.unload(TMB::dynlib(file.path("data", "test")))
```
