---
title: Variograms and the Leukaemia example
author: Kelli Faye Johnson
output: pdf_document
date: "`r format(Sys.time(), '%d %B, %Y')`"
bibliography: "c:/users/kelli/Google Drive/references/references.bib"
csl: "c:/users/kelli/Google Drive/references/ices-journal-of-marine-science.csl"
---

# Introduction

Template Model Builder (TMB), a statistical framework, used for the
estimation of parameters, including random effects was used to fit
data also fit by @lindgren2011, who used the `INLA` package in **R**.



```{r setup, echo = FALSE, results = FALSE}
library(geoR)
library(gstat)
require(INLA)
require(lattice)
require(rgl)
data(Leuk)
cpp <- "Leuk"
# May not be mac compatible
savedin <- gsub("\\\\", ,"/", tempdir())
```

```{r sinkcpp, echo = FALSE, results = FALSE}
sink(file = file.path(savedin, paste0(cpp, ".cpp")))
cat(
"// Illustration SPDE/INLA approach to spatial modelling via Matern correlation function
// Leukemia example from Lindgren et al 2011, JRSS-B
// http://www.r-inla.org/examples/case-studies/lindgren-rue-and-lindstrom-rss-paper-2011

#include <TMB.hpp>

template<class Type>
Type objective_function<Type>::operator() ()
{
  using namespace R_inla;
  using namespace density;
  using namespace Eigen;

  DATA_VECTOR(time);
  DATA_IVECTOR(notcens);
  DATA_IVECTOR(meshidxloc);
  DATA_MATRIX(X);

  DATA_STRUCT(spde,spde_t);

  PARAMETER_VECTOR(beta);
  PARAMETER(log_tau);
  PARAMETER(log_kappa);
  PARAMETER(log_omega);
  PARAMETER_VECTOR(x);

  Type tau = exp(log_tau);
  Type kappa = exp(log_kappa);
  Type omega = exp(log_omega);  // Parameter of Weibull distribution

  Type nll = 0.0;

  SparseMatrix<Type> Q = Q_spde(spde,kappa);

  nll = GMRF(Q)(x);                              // Negative log likelihood

  vector<Type> Xbeta = X*beta;
  for(int i=0; i<time.size(); i++){
    Type eta = Xbeta(i) + x(meshidxloc(i))/tau;
    Type lambda = exp(eta);
    Type t_omega = pow(time(i),omega);
    Type S = exp(-lambda*t_omega);               // Survival function
    Type f = lambda*omega*t_omega/time(i)*S;     // Weibull density

    // Likelihood contribution depends on truncation status
    if(notcens(i))
      nll -= log(f);
    else
      nll -= log(S);
  }

  double nu = 1.0;            // nu = alpha-d/2 = 2-1 by eqn (2) in Lindgren
  Type rho = sqrt(8*nu)/kappa;  // Distance at which correlation has dropped to 0.1 (p.  4 in Lindgren)

  REPORT(beta);
  REPORT(log_tau);
  REPORT(log_kappa);
  REPORT(log_omega);
  REPORT(x);
  ADREPORT(rho);

  return nll;
}
")
sink()

compile(file.path(savedin, paste0(cpp, ".cpp")))
dyn.load(dynlib(file.path(savedin, cpp)))
```

```{r tmb}
loc <- cbind(Leuk$xcoord, Leuk$ycoord)
mesh <- inla.mesh.2d(
  loc = cbind(Leuk$xcoord, Leuk$ycoord),
  boundary = list(inla.nonconvex.hull(loc, convex = 0.05),
    inla.nonconvex.hull(loc, convex = 0.25)),
  min.angle = 24,
  max.edge = c(0.05, 0.2),
  cutoff = 0.005,
  plot.delay = 0.5
  )

# Fixed effects part of model
data <- list(time = Leuk$time, notcens = Leuk$cens,
  meshidxloc = mesh$idx$loc - 1,
  X = as.matrix(model.matrix(~ 1 + sex + age + wbc + tpi, data = Leuk)))

# SPDE part: builds 3 components of Q (precision matrix)
# Encapsulation of 3 matrices
data$spde <- (inla.spde2.matern(mesh, alpha=2)$param.inla)[c("M0","M1","M2")]
# Number of points in mesh (including supporting points)
n_s = nrow(data$spde$M0)

parameters <- list(
  beta = c(-5.0, 0, 0, 0, 0),
  log_tau = -2.0,
  log_kappa = 2.5,
  log_omega = -1,
  x = rep(0.0, n_s))

# Phase 1: Fit non-spatial part first to get good starting values for fixed effects
obj <- MakeADFun(data,
  parameters,
  map = list(log_tau = as.factor(NA), log_kappa = as.factor(NA), x = factor(rep(NA,n_s))),
  DLL = cpp,
  silent = TRUE)
opt1 <- nlminb(obj$par, obj$fn, obj$gr)

# Modify starting values after phase 1
parameters <- list(
  beta = opt1$par[1:5],
  log_tau = -2.0,
  log_kappa = 2.5,
  log_omega = opt1$par["log_omega"],
  x = rep(0.0,n_s))

# Phase 2: Include spatial part. Use starting values from phase 1
obj <- MakeADFun(data,parameters, random = "x", DLL = cpp, silent = TRUE)
opt <- nlminb(obj$par,obj$fn,obj$gr,
  lower = c(-7, -1, -1, -1, -1, -3.0, 2.0, log(0.1)),
  upper = c(-4, 1, 1, 1, 1, -1.0, 3.0, log(10.0)))

# Calculate standard deviations, and extract rho
Rep <- sdreport(obj)
```

# Variogram estimation

## Estimate parameters by MLE
Use the `geoR::likfit` function to estimate parameters of the variogram.

```{r variogram}
data_var <- cbind(mesh$loc[, 1:2], obj$report()$x)
colnames(data_var) <- c("x", "y", "space")

v0 <- gstat::variogram(space ~ 1, data_var)
fit.variogram(v0, vgm(c("Exp", "Mat", "Sph")))

vg <- variog(data = obj$report()$x,
  coords = mesh$loc, uvec = seq(0, 0.2, len = 100))

variofit1 <- variofit(
  vario = vg, cov.model = "matern",
  ini.cov.pars = c(0.0, 0.1),
  fix.nugget = TRUE,
  nugget = 0.0, weight = "equal")

geodata <- geoR::as.geodata(cbind(mesh$loc, obj$report()$x))
# MLE
## trend is the mean function we are going to use. "cte" is the only intercept model
ml1 <- geoR::likfit(geodata,
  fix.nugget = FALSE,
  nugget = 0,
  cov.model = "matern",
  ini = c(0.1, 0.1),
  # ini = variofit1$cov.pars,
  trend = "cte")

plot(vg)
lines.variomodel(x = bins, cov.model = "matern",
  cov.pars = ml1$cov.pars, nugget = ml1$nugget, lty = 3)
legend("bottomright", c("OLS","MLE"), inset=0.05, lty=1:3)
```

```{r lindgrenetal2011, echo = TRUE, eval = FALSE}
## Code for Lindgren, Rue, Lindstrom (2011), partially updated to newer R-INLA
## Note: Will not reproduce exactly the same results as in the paper.

# Read data
temp <- tempfile()
download.file("http://www.math.ntnu.no/inla/r-inla.org/case-studies/Lindgren-Rue-Lindstrom-rss-2011/leuk-demo.zip",
  temp)
source(unz(temp, "leuk-demo-new2/utils.R"))
source(unz(temp, "leuk-demo-new2/utils.R"))
map <- read.table(unz(temp, "leuk-demo-new2/Leuk.map"))
Leuk$id = 1:dim(Leuk)[1]

## Build triangular mesh:
mesh = (inla.mesh.2d(
  ## Data locations to use as location seeds:
  loc=cbind(Leuk$xcoord, Leuk$ycoord),
  ## Encapsulate data region:
  boundary=list(inla.nonconvex.hull(cbind(Leuk$xcoord, Leuk$ycoord), convex=0.05),
    inla.nonconvex.hull(cbind(Leuk$xcoord, Leuk$ycoord), convex=0.25)),
  ## Refined triangulation,
  ## minimal angles >=26 degrees,
  ## interior maximal edge lengths 0.05,
  ## exterior maximal edge lengths 0.2,
  ## don't add input points closer than 0.05:
  min.angle=24,
  max.edge=c(0.05, 0.2),
  cutoff=0.005,
  ## Set to >=0 for visual (no effect Windows):
  plot.delay=0.5
  ))
## Store the data-->vertex mapping:
## ( mesh$idx$loc[k] is the mesh vertex for data location nr. k )
Leuk$spatial = mesh$idx$loc

## Create the SPDE/GMRF model, (kappa^2-Delta)(tau x) = W:
spde = inla.spde2.matern(mesh, alpha=2)

## Build the GLM model:
formula = inla.surv(Leuk$time, Leuk$cens) ~ 1 + sex + age + wbc + tpi +
    ## Add the spatial effect model:
    f(spatial, model=spde)

## Run INLA:
r  = (inla(formula, family="weibull",
           data = Leuk,
           ## Prior specification:
           control.family = list(param=c(0.05,0.1)),
           ## Reasonable starting point for the optimisation:
           control.mode = list(theta=c(-0.5,-2,2),restart=TRUE),
           ## We don't need the marginals:
           control.compute = list(return.marginals=FALSE),
           ## We don't need to overoptimise:
           control.inla=list(tolerance=1e-5),
           ## Verbose output:
           verbose=TRUE
           ))

## Extract the SPDE parameters:
tau <- exp(r$summary.hyperpar[2,"mean"])
kappa <- exp(r$summary.hyperpar[3,"mean"])
range <- sqrt(8) / kappa

## Get the precision matrix:
Q = inla.spde2.precision(spde, theta=log(c(tau, kappa)))


## Get a random sample (not used here),
## and the index reordering,
## using an undocumented ad hoc function:
reo=inla.qreordering(Q, reordering="metis")
## Need to invert the indexing:
neworder = reo$reordering
neworder[neworder] = 1:length(neworder)
## Reorder the matrix:
Q.reordered = Q[ neworder,neworder ]


## Reference point for covariance/correlation comparisons:
ref.s = (which.min((mesh$loc[,1]-mean(range(mesh$loc[,1])))^2 +
                   (mesh$loc[,2]-mean(range(mesh$loc[,2])))^2))


## Calculate covariances (S) and correlations (SS):
S = solve(Q)
SS = diag(1/sqrt(diag(S))) %*% S %*% diag(1/sqrt(diag(S)))
D = as.matrix(dist(mesh$loc))

## Theoretical Matern correlations and covariances:
dd = (0:1000)/1000
SS.theory = (dd*kappa)*besselK(dd*kappa,1)
SS.theory[1] = 1
S.theory = SS.theory/(4*pi*kappa^2)/tau^2


###########################
## Prepare for plotting:

## Calculate mapping between triangulation vertices and grid points:
## Resolution for gridded output was dims=c(100,100) in the paper.
proj = inla.mesh.projector(mesh, dims=c(200,200))

## Construct greyscale palette function:
my.grey.palette = function (n,...) { return (grey.colors(n,0.05,0.95,...))}
## Use it:
my.palette = my.grey.palette

## Construct map data appropriate for easy plotting:
mm = calc.map(map)


#####################
## Plot results:

## Compare correlations:
dev.new()
plot(D[ref.s,],SS[ref.s,],type="p",pch=20,
     xlab="Distance",ylab="Correlation",
     xlim=c(0.0,1.0),ylim=c(-0.005,1.005))
lines(dd,SS.theory,type="l",
      col=rgb(0.5,0.5,0.5),lwd=2)
ka <- exp(Rep$par.fixed["log_kappa"]); ta <- tau
test = (dd*ka)*besselK(dd*ka,1)
test[1] = 1
lines(dd, test, col = "blue")

## Don't plot the full precision pattern; reduce the dimension first:
Q.coarse = sparse.pattern.coarsen(Q.reordered,4)
## Plot the reordered precision pattern:
dev.new()
image(as.matrix(Q.coarse)>0,
      col=grey.colors(2,start=1,end=0),
      ylim=c(1,0), axes=FALSE, xlim=c(0,1))
box()



## Map resulting posterior mean field to a grid:
plotdata = inla.mesh.project(proj, r$summary.random$spatial[,"mean"])
## Plot PM contours:
dev.new()
bbb = (levelplot(row.values=proj$x, column.values=proj$y, x=plotdata,
                 mm=mm, panel=levelplotmap,
                 col.regions=my.palette,
                 xlim=range(proj$x), ylim=range(proj$y), aspect="iso",
                 contour=TRUE, cuts=11, labels=FALSE, pretty=TRUE,
                 xlab="Easting",ylab="Northing"))
print(bbb)


## Map resulting posterior standard deviation field to a grid:
plotdata = inla.mesh.project(proj, r$summary.random$spatial[,"sd"])
## Plot std.dev. contours:
dev.new()
bbb = (levelplot(row.values=proj$x, column.values=proj$y, x=plotdata,
                 mm=mm, panel=levelplotmap,
                 col.regions=my.palette,
                 xlim=range(proj$x), ylim=range(proj$y), aspect="iso",
                 contour=TRUE, cuts=11, labels=FALSE, pretty=TRUE,
                 xlab="Easting",ylab="Northing"))
print(bbb)


## Plot data points+map:
rgl.win1 = open3d()
plot.rgl.map(map, color = "black", lwd=2, zoffset=0.1)
plot.fake.points(cbind(Leuk$xcoord, Leuk$ycoord,0.1), radius=0.005,
                 color="black",specular="black")
plot(mesh, rgl=TRUE, add=TRUE,
     lwd=1, draw.vertices=FALSE, draw.edges=FALSE,
     col="white");
view3d(0,0,fov=0,zoom=0.8)

## Plot triangulation+map
rgl.win2 = open3d()
plot.rgl.map(map, color = "black", lwd=1, zoffset=0.1)
plot(mesh, rgl=TRUE, add=TRUE,
     draw.vertices=FALSE, draw.edges=TRUE,
     edge.color=rgb(0.6,0.6,0.6), lwd=1, col="white")
view3d(0,0,fov=0,zoom=0.8)
```

# References
@article {lindgren2011,
author = {Lindgren, F. and Rue, H. and Lindstr\"{o}m, J.},
title = {An explicit link between Gaussian fields and Gaussian Markov random fields: the stochastic partial differential equation approach},
journal = {Journal of the Royal Statistical Society: Series B (Statistical Methodology)},
year = {2011},
volume = {73},
number = {4},
pages = {423-498},
url = {http://dx.doi.org/10.1111/j.1467-9868.2011.00777.x},
doi = {10.1111/j.1467-9868.2011.00777.x},
keywords = {Approximate Bayesian inference, Covariance functions, Gaussian fields, Gaussian Markov random fields, Latent Gaussian models, Sparse matrices, Stochastic partial differential equations},
}

@Article{henderson2002,
  Author = {Henderson, R. and Shimakura, S. and Gorst, D.},
  Title = {Modelling spatial variation in leukemia survival data},
  Journal = {Journal of the American Statistical Association},
  Year = {2002},
  Volume = {97},
  Pages = {965-972},
  }
